The mTOR pathway genes mTOR, Rheb, Depdc5, Pten, and Tsc1 have convergent and divergent impacts on cortical neuron development and function

Version 1

Abstract

Abstract

Brain somatic mutations in various components of the mTOR complex 1 (mTORC1) pathway have emerged as major causes of focal malformations of cortical development and intractable epilepsy. While these distinct gene mutations converge on excessive mTORC1 signaling and lead to common clinical manifestations, it remains unclear whether they cause similar cellular and synaptic disruptions underlying cortical network hyperexcitability. Here, we show that in utero activation of the mTORC1 activators, Rheb or mTOR, or biallelic inactivation of the mTORC1 repressors, Depdc5, Tsc1, or Pten in mouse medial prefrontal cortex leads to shared alterations in pyramidal neuron morphology, positioning, and membrane excitability but different changes in excitatory synaptic transmission. Our findings suggest that, despite converging on mTORC1 signaling, mutations in different mTORC1 pathway genes differentially impact cortical excitatory synaptic activity, which may confer gene-specific mechanisms of hyperexcitability and responses to therapeutic intervention.

Introduction

Focal malformations of cortical development (FMCDs), including focal cortical dysplasia type II (FCDII) and hemimegalencephaly (HME), are the most common causes of intractable epilepsy in children1,2. These disorders are characterized by abnormal brain cytoarchitecture with cortical overgrowth, mislamination, and cellular anomalies, and are often associated with developmental delay and intellectual disability3. Early immunohistochemical studies identified hyperactivation of the mechanistic target of rapamycin complex 1 (mTORC1) signaling pathway in resected brain tissue from individuals with FCDII and HME, leading to the classification of these disorders as “mTORopathies”4. More recently, somatic mutations in numerous regulators of the mTORC1 pathway were identified as the genetic causes of FCDII and HME5–7. Accumulating evidence shows that these mutations occur in a subset of dorsal telencephalic progenitor cells that give rise to excitatory neurons during fetal development, resulting in brain somatic mosaicism8,9. These genetic findings provide opportunities for gene therapy approaches targeting the mTORC1 pathway for FCDII and HME.

mTORC1 is an evolutionarily conserved protein complex that promotes cell growth and differentiation through the regulation of protein synthesis, metabolism, and autophagy10. mTORC1 is composed of mTOR, a serine/threonine kinase that exerts the complex’s catalytic function, and several companion proteins. Activation of mTORC1 is controlled by a well-described upstream cascade involving multiple protein regulators11. Stimulation by growth factors activates phosphoinositide 3-kinase (PI3K), which leads to the activation of phosphoinositide-dependent kinase 1 (PDK1) and subsequent phosphorylation and activation of AKT. Activated AKT phosphorylates and inactivates tuberous sclerosis complex 1/2 (TSC1/2 complex; consisting of TSC1 and TSC2 proteins), which releases the brake on Ras homolog enriched in brain (RHEB), a GTP-binding protein that directly activates mTORC1. This mTORC1-activating cascade is negatively controlled by the phosphatase and tensin homolog (PTEN) protein, which inhibits PI3K activation of PDK1. Additionally, mTORC1 signaling is regulated by a separate nutrient-sensing GAP Activity Towards Rags 1 (GATOR1) complex, which consists of DEP domain containing 5 (DEPDC5) and the nitrogen permease regulator 2-like (NPRL2) and 3-like (NPRL3) proteins12. The GATOR1 complex serves as a negative regulator of mTORC1 that inhibits mTORC1 at low amino acid levels.

Pathogenic mutations in numerous regulators that activate mTORC1 signaling, including PIK3CA, PTEN, AKT3, TSC1, TSC2, MTOR, RHEB, DEPDC5, NPRL2, and NPRL3, have been identified in HME and FCDII9. Genetic targeting of these genes in mouse models consistently recapitulates the epilepsy phenotype, supporting an important role for these genes in seizure generation13,14. However, while all these genes impinge on the mTORC1 pathway, many of them also participate in mTORC1-independent functions through parallel signaling pathways13,15–19, and it is unclear whether mutations affecting different mTORC1 pathway genes lead to cortical hyperexcitability through common neural mechanisms. In this study, we examined how disruption of five distinct mTORC1 pathway genes, Rheb, mTOR, Depdc5, Pten, and Tsc1, individually impact pyramidal neuron development and electrophysiological function in the mouse medial prefrontal cortex (mPFC). Collectively, we found that activation of the mTORC1 activators, Rheb and mTOR, or inactivation of the mTORC1 repressors, Depdc5, Tsc1, and Pten, largely leads to similar alterations in neuron morphology and membrane excitability but differentially impacts excitatory synaptic activity. The latter has implications for cortical network function and seizure vulnerability and may affect how individuals with different genotypes respond to targeted therapeutics.

Results

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOleads to varying magnitudes of neuronal enlargement and mispositioning in the cortex

To model somatic gain-of-function mutations in the mTORC1 activators, we individually expressed plasmids encoding RhebY35L or mTORS2215Y, respectively, in select mouse neural progenitor cells during late corticogenesis via in utero electroporation (IUE) (Fig. 1a, b). These pathogenic variants were previously detected in brain tissue from children with FCDII and HME associated with intractable seizures20–25. We specifically targeted a subset of late-born progenitor cells that generate excitatory pyramidal neurons destined to layer 2/3 in the medial prefrontal cortex (mPFC) to mimic frontal lobe somatic mutations and the genetic mosaicism characteristic of these lesions. To model the somatic loss-of-function mutations in mTORC1 repressors, we expressed plasmids encoding previously validated CRISPR/Cas9 guide RNAs against Depdc526, Tsc127, or Pten28 to individually knockout (KO) the respective genes using the same IUE approach (Fig. 1a, b). As a control for the activating mutations, we expressed plasmids encoding fluorescent proteins under the same CAG promoter. As a control for the CRISPR/Cas9-mediated knockouts, we used an empty CRISPR/Cas9 vector containing no guide RNA sequences. To verify that the expression of these plasmids leads to mTORC1 hyperactivation, we assessed the phosphorylated levels of S6 (i.e., p-S6), a downstream substrate of mTORC1, using immunostaining in brain sections from postnatal day (P) 28-43 mice. As expected, we found that expression of the RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO plasmids all led to significantly increased p-S6 staining intensity, supporting that the expression of each of these plasmids leads to increased mTORC1 signaling (Fig. 1c, d, Table 1, Fig. S1a).

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOleads to varying magnitudes of neuronal enlargement and mispositioning in the cortex.

(a) Diagram of the PI3K-mTORC1 pathway. Activation of mTORC1 signaling is controlled by positive (blue) and negative (purple) regulators within the pathway. (b) Diagram of overall experimental timeline and approach. IUE was performed at E15.5. A cohort of animals was used for patch clamp recording at P26-51 and another cohort was used for histology at P28-43. (c) Representative images of tdTomato+ cells (red) and p-S6 staining (green) in mouse mPFC at P28-43. (d) Quantification of p-S6 staining intensity (normalized to the mean control) in tdTomato+ neurons. (e) Quantification of tdTomato+ neuron soma size. (f) Representative images of tdTomato+ neuron (red) placement and distribution in coronal mPFC sections. Red square on the diagram denotes the imaged area for all groups. CC, corpus callosum. (g) Quantification of tdTomato+ neuron placement in layer 2/3. Left diagram depicts the approach for analysis: the total number of tdTomato+ neurons within layer 2/3 (white square) was counted and expressed as a % of total neurons in the imaged area. Right bar graphs show the quantification. (h) Quantification of tdTomato+ neuron distribution across cortical layers. Left diagram depicts the approach for analysis: the imaged area was divided into 10 equal-sized bins across the cortex, spanning the corpus callosum to the pial surface (white grids); the total number of tdTomato+ neurons within each bin was counted and expressed as a % of total neurons in the imaged area. Right bar graphs show the quantification. For graphs d, e: n = 4-8 mice per group, with 6-15 cells analyzed per animal. For graphs g, h: n = 3-7 mice per group, with 1 brain section analyzed per animal. Statistical comparisons were performed using (d, e) nested one-way ANOVA (fitted to a mixed-effects model) to account for correlated data within individual animals, (g) one-way ANOVA, or (h) two-way repeated measures ANOVA. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001. All data are reported as the mean of all neurons or brain sections in each group ± SEM.

Considering the cytoarchitectural abnormalities associated with mTORC1 hyperactivation, we compared neuron soma size and positioning in the cortex following in utero activation of Rheb and mTOR and inactivation of Depdc5, Pten, and Tsc1 in P28-43 mice. While all experimental conditions led to increased neuron soma size, the magnitude of the enlargement was dependent on the specific gene that was targeted (Fig. 1c, e, Table 1. Fig. S1b). In particular, expression of RhebY35L and mTORS2215Y led to the largest soma size changes, with a >3-fold increase from control. Expression of PtenKO led to a similarly large increase of 2.7-fold, while expression of Dedpdc5KOand Tsc1KO led to a 1.7 and 2.1-fold increase, respectively. The increase in the PtenKO condition was significantly higher than both the Dedpdc5KO and Tsc1KO conditions, and the increase in the Tsc1KO condition was significantly higher than the Dedpdc5KO condition. Although the above analysis was performed at P28-43, the enlargement of neuron soma sizes was already detected by P7-9 (Fig. S2a-c, Table S1). In terms of neuronal positioning, all experimental conditions, except for PtenKO, resulted in neuron misplacement (Fig. 1f-h, Table 1). The RhebY35L and mTORS2215Yconditions led to the most severe phenotype: ∼70-80% of the neurons were misplaced outside of layer 2/3 (Fig. 1g), with the mispositioned neurons evenly scattered across the deeper layers (Fig. 1h). For the Dedpdc5KO and Tsc1KO conditions, ∼45-60% of the neurons were misplaced outside of layer 2/3 (Fig. 1g), with the mispositioned neurons found closer to layer 2/3 (Fig. 1h). Taken together, these studies show that the expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO leads to varying magnitudes of neuronal enlargement and mispositioning in the cortex. Of note, PtenKO neurons, despite exhibiting a 2.7-fold increase in neuron soma size, were mostly correctly positioned in layer 2/3. These findings suggest that while all experimental conditions lead to increased soma size, not all lead to neuron mispositioning, suggesting defective migration and the subsequent impact on neuron positioning occur independently of cell size.

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOuniversally leads to decreased depolarization-induced excitability, but only RhebY35L, mTORS2215Y, and Tsc1KOexpression leads to depolarized resting membrane potential

To elucidate the contribution of each experimental condition to the function of cortical neurons, we obtained whole-cell patch clamp recordings of layer 2/3 pyramidal neurons at P26-P51 (Fig. 2a). The RhebY35Land mTORS2215Yconditions were compared to a control group expressing fluorescent proteins under the same CAG promoter. The Dedpdc5KO, PtenKO, and Tsc1KO conditions were compared to a CRISPR/Cas9 empty vector control. Recordings of control and experimental conditions were alternated to match the animal ages.

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOuniversally leads to decreased depolarization-induced excitability, but only RhebY35L, mTORS2215Y, and Tsc1KO expression leads to depolarized RMPs.

(a) Diagram of experimental approach for whole-cell patch clamp recording. Recordings were performed in layer 2/3 pyramidal neurons in acute coronal slices from P26-51 mice, expressing either control, RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, or Tsc1KO plasmids. (b-d) Graphs of membrane capacitance, resting membrane conductance, and RMP. (e) Representative traces of the AP firing response to a 300 pA depolarizing current injection. (f) Input-output curves showing the mean number of APs fired in response to 500 ms-long depolarizing current steps from 0 to 400 pA. Arrows point to initial spike doublets. (g-j) Graphs of rheobase, AP threshold, AP peak amplitude, and AP half-width. For all graphs: n = 5-10 mice per group, with 22-50 cells analyzed per animal. Statistical comparisons were performed using (b-d, g-j) nested one-way ANOVA (fitted to a mixed-effects model) to account for correlated data within individual animals or (f) mixed-effects ANOVA accounting for repeated measures. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001, for the input-output curves in f: *p<0.05 (vs. control), #<0.05 (vs. Dedpdc5KO), Ɏp<0.05 (vs. Tsc1KO). All data are reported as the mean of all neurons in each group ± SEM.

Consistent with the changes in soma size (Fig. 1c, e), recorded neurons displayed increased membrane capacitance in all experimental conditions (Fig. 2b, Table 1, Fig. S3a). Neurons expressing the mTORS2215Yvariant had a larger capacitance increase than those expressing the RhebY35L variant. PtenKO and Tsc1KO neurons had similar increases in capacitance that were both larger than that of the Depdc5KO neurons. All neurons across the experimental conditions also had increased resting membrane conductance in a pattern that followed that of the capacitance (Fig. 2c, Table 1, Fig. S3b). However, while RhebY35L, mTORS2215Y, and Tsc1KOexpression led to depolarized resting membrane potential (RMP), Dedpdc5KOand PtenKO expression did not significantly affect the RMP (Fig. 2d, Table 1, Fig. S3c). To assess whether these changes impacted neuron intrinsic excitability, we examined the action potential (AP) firing response to depolarizing current injections. For all experimental conditions, neurons fired fewer APs for current injections above 100 pA compared to the respective control neurons (Fig. 2e, f, Table 1). This decrease in intrinsic excitability is reflected in the increased rheobase (i.e., the minimum current required to induce an AP) in all experimental conditions, with the mTORS2215Y and RhebY35Lconditions leading to the largest rheobase increases (Fig. 2g, Table 1, Fig. S3d). Collectively, these findings indicate that RhebY35L, mTORS2215Y, and Tsc1KO neurons display a decreased ability to generate APs upon depolarization despite having depolarized RMPs. In terms of firing pattern, neurons in all groups displayed a regular-spiking pattern with spike-frequency adaptation (Fig. 2e). However, while an initial spike doublet (with an interspike interval of 10-15 ms) was observed in the majority of control neurons, consistent with the expected firing pattern for layer 2/3 mPFC pyramidal neurons29, this was observed in a smaller fraction of neurons in the experimental conditions [number of neurons displaying spike doublets: RhebY35L: 4/23 (17%), mTORS2215Y: 7/28 (25%), Dedpdc5KO: 11/27 (41%), PtenKO: 21/28 (75%) - notably, 6 of the 21 doublets are parts of triplets, Tsc1KO: 7/28 (25%), control: 24/26 (92%)]. There were no differences in the AP threshold across the conditions (Fig. 2h, Table 1, Fig. S3e). The AP peak amplitude was decreased in the mTORS2215Yand Tsc1KO conditions (Fig. 2i, Table 1, Fig. S3f), while the AP half-width was decreased in the RhebY35L, mTORS2215Y, and PtenKO conditions (Fig. 2j, Table 1, Fig. S3g). Taken together, these findings show that various genetic conditions that activate the mTORC1 pathway universally lead to decreased depolarization-induced excitability in layer 2/3 pyramidal neurons, with gene-dependent changes in RMP and several AP properties.

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOleads to the abnormal presence of HCN4 channels with variations in functional expression

We recently reported that neurons expressing RhebS16H, an mTOR activating variant of Rheb, display abnormal expression of HCN4 channels30,31. These channels give rise to a hyperpolarization-activated cation current (Ih) that is normally absent in layer 2/3 pyramidal neurons30,31. The aberrant Ih, which has implications for neuronal excitability, preceded seizure onset and was detected by P8-12 in mice30. Rapamycin treatment starting at P1 and expression of constitutive active 4E-BP1, a translational repressor downstream of mTORC1, prevented and alleviated the aberrant HCN4 channel expression, respectively30,31. These findings suggest that the abnormal HCN4 channel expression is mTORC1-dependent. Given that all the experimental conditions in the present study led to increased mTORC1 activity, we investigated whether they also resulted in abnormal HCN4 channel expression. Immunostaining for HCN4 channels using previously validated antibodies30 in brain sections from P28-43 mice revealed significantly increased HCN4 staining intensity in the electroporated neurons in all experimental conditions compared to the respective controls, which exhibited no HCN4 staining (Fig 3a, b, Table 1, Fig. S4a). The increased staining was evident in the ipsilateral cortex containing mTORS2215Yelectroporated neurons and absent in the non-electroporated contralateral cortex (Fig. S5a). These data indicate the presence of aberrant HCN4 channel expression following RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, or Tsc1KO expression in layer 2/3 pyramidal neurons.

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOleads to the abnormal presence of HCN4 channels with variations in functional expression.

(a) Representative images of tdTomato+ cells (red) and HCN4 staining (green) in mouse mPFC at P28-43. (b) Quantification of HCN4 intensity (normalized to the mean control) in tdTomato+ neurons. (c) Representative current traces in response to a series of 3 s-long hyperpolarizing voltage steps from -120 to -40 mV, with a holding potential of -70 mV. Current traces from the -40 and -50 mV steps were not included due to contamination from unclamped Na+ spikes. Orange lines denote the current traces at -90 mV. (d) IV curve obtained from Iss amplitudes. (e) ΔIV curve obtained from Ih amplitudes (i.e., ΔI, where ΔI=Iss – Iinst). (f) Graphs of Ih amplitudes at –90 mV. (g) Representative current traces in response to a series of 3-s long hyperpolarizing voltage steps from -110 mV to -50 mV in the mTORS2215Y condition pre- and post-zatebradine application. Orange lines denote the current traces at -90 mV. (h) IV curve obtained from Iss amplitudes in the mTORS2215Y condition pre- and post-zatebradine application. (i) ΔIV curve obtained from Ih amplitudes (i.e., ΔI) in the mTORS2215Y condition pre- and post-zatebradine application. Arrow points to the post-zatebradine Ih amplitude at -90 mV. (j) Representative traces of the zatebradine-sensitive current obtained after subtraction of the post-from the pre-zatebradine current traces in response to -110 mV to -50 mV voltage steps. Orange lines denote the current traces at -90 mV. (k) IV curve of the zatebradine-sensitive current obtained after subtraction of the post-from the pre-zatebradine current traces. (l) Graph of RMP in the control and mTORS2215Y conditions pre- and post-zatebradine application. Connecting lines denote paired values from the same neuron. For graph b: n = 4-8 mice per group, with 4-15 cells analyzed per animal. For graphs d, e, f: n = 5-10 mice per group, with 24-47 cells analyzed per animal. For graphs h, i, k, l: n = 4-6 neurons (paired). Statistical comparisons were performed using (b, f) nested ANOVA (fitted to a mixed-effects model) to account for correlated data within individual animals, (d, e) mixed-effects ANOVA accounting for repeated measures, or (h, i, l) two-way repeated measures ANOVA. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ****p<0.0001, for the IV curves in d, e, h, i: *p<0.05 (vs. control), #p<0.05 (vs. RhebY35L or Dedpdc5KO), Ɏp<0.05 (vs. PtenKO). All data are reported as the mean of all neurons in each group ± SEM.

To examine the functional impacts of the aberrant HCN4 channel expression, we examined Ih amplitudes in the various experimental conditions. To evoke Ih, we applied a series of 3 s-long hyperpolarizing voltage steps from -120 mV to -40 mV. Consistent with previous findings in RhebS16H neurons30,31, hyperpolarizing voltage pulses elicited significantly larger inward currents in all experimental conditions compared to their respective controls (Fig. 3c, d, Table 1). The mTORS2215Y condition displayed larger inward currents than the RhebY35L condition, while the PtenKO condition displayed the largest inward currents compared to the Depdc5KO and Tsc1KO conditions (Fig. 3d). These data were proportional to the changes in neuron soma size (Fig. 1e, 2b). The inward currents in RhebS16H neurons are thought to result from the activation of both inward-rectifier K+ (Kir) channels and HCN channels30. Kir-mediated currents activate fast whereas HCN-mediated currents, i.e., Ih, activate slowly during hyperpolarizing steps; therefore, to assess Ih amplitudes, we measured the difference between the instantaneous and steady-state currents at the beginning and end of the voltage pulses, respectively (i.e., ΔI)32. The resulting ΔI-voltage (V) curve revealed significantly larger Ih amplitudes in all experimental conditions (Fig. 3e, Table 1). To further isolate the Ih from Kir-mediated currents, we measured Ih amplitudes at -90 mV, near the reversal potential of Kir channels to eliminate Kir-mediated current contamination. At -90 mV, Ih amplitudes were significantly higher in the mTORS2215Y and Tsc1KO conditions compared to controls (Fig. 3f, Table 1, Fig. S4b). Of note, although the Depdc5KO and PtenKO conditions did not display a significant increase in Ih amplitudes at -90 mV, 1 out of 28 Depdc5KO neurons and 4 out of 27 PtenKO neurons had Ih amplitudes that were 2-fold greater than the mean Ih amplitude of the Tsc1KO condition. 6 out of 24 RhebY35L neurons also had Ih amplitudes 2-fold greater than this value (Fig. 3f). These data suggest that Ih currents are present in a subset of RhebY35L, Depdc5KO, and PtenKO neurons, and most mTORS2215Y and Tsc1KO neurons.

Considering that the mTORS2215Y condition led to the largest Ih, we examined the impact of the selective HCN channel blocker zatebradine on hyperpolarization-induced inward currents in mTORS2215Y neurons to further confirm the identity of ΔI as Ih. Application of 40 µM zatebradine reduced the overall inward currents (Fig. 3g, h, Table 1) and ΔI (Fig. 3i, Table 1). At -90 mV, ΔI was significantly decreased from -167.7 ± 54.2 pA to 0.75 ± 8.2 pA (± SD) (Fig. 3i, arrow). Subtraction of the post-from the pre-zatebradine current traces isolated zatebradine-sensitive inward currents which reversed near -50 mV, as previously reported for HCN4 channel reversal potentials33 (Fig. 3j, k, Table 1). These experiments verified the identity of ΔI as Ih. In comparison, application of the Kir channel blocker barium chloride (BaCl2) substantially reduced the overall inward currents but had no effects on ΔI (i.e., Ih) in Tsc1KOneurons (Fig. S6a-i). Consistent with the function of Ih in maintaining RMP at depolarized levels34–38, application of zatebradine hyperpolarized RMP in mTORS2215Y neurons but did not affect the RMP of control neurons that exhibited no Ih (Fig. 3l, Table 1). Collectively, these findings suggest that RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO expression in layer 2/3 pyramidal neurons lead to the abnormal presence of HCN4 channels with variations in functional expression.

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOleads to different impacts on excitatory synaptic activity

As part of examining neuron excitability, we recorded spontaneous excitatory postsynaptic currents (sEPSCs) in all the gene conditions. The frequency of sEPSCs was unchanged in all experimental conditions except for the Tsc1KO condition, in which the frequency was significantly increased (Fig. 4a, b, Table 1, Fig. S7a). Unlike the other experimental conditions, the RhebY35Lcondition displayed a slight decrease in sEPSC frequency, consistent with previous findings in RhebS16H neurons; however, this did not reach statistical significance (Fig. 4b). The amplitude of sEPSCs was larger in the RhebY35L, mTORS2215Y, and PtenKO conditions (Fig. 4a, c, Table 1, Fig. S7b). Although the amplitudes were slightly larger in the Depdc5KO and Tsc1KO conditions, these changes were not significant (Fig. 4c). Thus, Tsc1KO neurons display increased sEPSC frequency but unchanged amplitude, while RhebY35L, mTORS2215Y,and PtenKO neurons display increased sEPSC amplitude but unchanged frequency. Finally, there was an increase in the sEPSC total charge in all experimental conditions except for the RhebY35L condition (Fig. 4d, Table 1, Fig. S7c). Collectively, these findings suggest all experimental conditions, except for RhebY35L, lead to increased synaptic excitability, with variable impact on sEPSC frequency and amplitude.

Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KOleads to different impacts on sEPSC properties.

(a) Representative sEPSC traces recorded at a holding voltage of -70 mV. Top and bottom traces are from the same neuron. (b-d) Graphs of sEPSC frequency, amplitude, and total charge. For all graphs: n = 5-9 mice per group, with 17-34 cells analyzed per animal. Statistical comparisons were performed using a nested ANOVA (fitted to a mixed effects model) to account for correlated data within individual animals. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001. All data are reported as the mean of all neurons in each group ± SEM.

Discussion

In this unprecedented comparison study, we examined the impacts of several distinct epilepsy-associated mTORC1 pathway gene mutations on cortical pyramidal neuron development and their electrophysiological properties and synaptic integration in the mouse mPFC. Through a combination of IUE to model the genetic mosaicism of FMCDs, histological analyses, and patch-clamp electrophysiological recordings, we found that activation of either Rheb or mTOR or biallelic inactivation Depdc5, Tsc1, or Pten, all of which increase mTORC1 activity, largely leads to similar alterations in neuron morphology and membrane excitability but differentially impacts excitatory synaptic transmission. These findings have significant implications for understanding the mechanisms leading to cortical hyperexcitability and seizures in mTORC1-driven FMCDs and highlight the utility of personalized medicine dictated by patient gene variants.

Several histological phenotypes associated with mTORC1 hyperactivation were anticipated and confirmed in our studies. We found increased neuron soma size across all gene conditions, consistent with previous reports24,26,27,39–42. Additionally, all conditions, except for PtenKO, resulted in neuronal mispositioning in the mPFC, with RhebY35L and mTORS2215Y conditions being the most severe. Interestingly, PtenKO neurons were correctly placed in layer 2/3, despite having one of the largest soma size increases, suggesting that cell positioning is independent of cell size. The lack of mispositioning of PtenKO neurons was surprising as it is thought that increased mTORC1 activity leads to neuronal misplacement, and aberrant migration of PtenKO neurons has been reported in the hippocampus43. Given that PTEN has a long half-life, with a reported range of 5 to 20 hours or more depending on the cell type44–51, it is possible that following knockout at E15, the decreases in existing protein levels lagged behind the time window to affect neuronal migration. However, TSC1 also has a long half-life of ∼22 hours52,53, and knockout of Tsc1 at the same time point was sufficient to affect neuronal positioning. Mispositioning of neurons following Pten knockout in E15 rats was reported by one study, but this was a very mild phenotype compared to findings in the other gene conditions39. Therefore, these data suggest that PTEN has gene-specific biological mechanisms that contribute to the lack of severe neuronal mispositioning in the developing cortex. mTORS2215Yexhibited the strongest phenotypes in terms of neuron size and positioning, which is perhaps not surprising as the mTOR protein itself forms the catalytic subunit of mTORC1. Depdc5KO exhibited the mildest changes in these parameters. Unlike PTEN, TSC1, RHEB, and mTOR, which regulate mTORC1 via the canonical PI3K-mTORC1 pathway in response to growth factor stimulation, DEPDC5 regulates mTORC1 via the GATOR1 complex in response to changes in cellular amino acid levels54. The different modes of mTORC1 regulation by DEPDC5 may contribute to the differences in the severity of the phenotypes.

To assess the impact of RHEB, mTOR, DEPDC5, PTEN, and TSC1 disruption on cortical neuron excitability, we examined the intrinsic biophysical and synaptic properties of layer 2/3 pyramidal neurons in the mPFC. We found that all gene manipulations led to increased membrane conductance but decreased AP firing upon membrane depolarization, consistent with previous reports in the somatosensory cortex26,39,42,55–58 and in the mPFC for the RhebS16Hcondition30,31. Neurons in these conditions required a larger depolarization to generate an AP (increased rheobase), likely due to their enlarged cell size. These findings seem to contradict the theory that seizure initiation originates from these neurons. However, inhibiting firing in cortical pyramidal neurons expressing the mTORC1-activating RhebS16H mutation via co-expression with Kir2.1 channels (to hyperpolarize neurons) has been shown to prevent seizures, supporting a cell-autonomous mechanism30. This discrepancy may be reconciled by the identification of abnormal HCN4 channel expression in the RhebS16H neurons, which was shown to contribute to increased excitability by depolarizing RMPs, bringing neurons closer to the AP threshold, and conferring firing sensitivity to cyclic AMP30. The abnormal HCN4 channel expression has been verified in resected human FCDII and HME tissue (of unknown genetic etiology)30. Here, we found that RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO cortical neurons also display aberrant HCN4 channel expression, which is consistent with the rapamycin- and 4E-BP1-dependent mode of expression in RhebS16H neurons30,31. The functional expression of these channels was variable, and the corresponding changes in Ih did not reach statistical significance for RhebY35L, Depdc5KO, and PtenKO neurons. Nonetheless, consistent with the function of HCN4 channels in maintaining the RMP at depolarized levels38, RhebY35L, mTORS2215Y, and Tsc1KO neurons exhibited depolarized RMPs. Interestingly, the RMP was unchanged in PtenKO and Depdc5KO neurons. The lack of RMP changes may be explained by the milder HCN4 phenotype or the presence of different ion channel complements in these neurons. In addition to HCN4 channels, it is thought that increased Kir channels, which are well-known to hyperpolarize RMPs, contribute to the enlarged inward currents in all the conditions we investigated. This was confirmed in the Tsc1KO neurons where application of BaCl2 to block Kir channels reduced the overall inward current and depolarized RMPs. Since PtenKO neurons have a larger overall inward current but a smaller Ih compared to Tsc1KO neurons, we postulate that PtenKO neurons have a higher expression of Kir channels neurons that could counteract the HCN4 channels’ depolarizing effect on RMP. The mechanisms and functional significance of the Kir channel increases in these neurons are yet to be validated. Nevertheless, our data suggest that abnormal HCN4 channel expression is a conserved mechanism across these gene conditions and warrant further investigation of HCN-mediated excitability in mTORC1-related epilepsy.

While membrane excitability was largely similar across all gene conditions, the excitatory synaptic properties were more variable. All conditions, except for RhebY35L, led to increased excitatory synaptic activity, with variable impact on sEPSC frequency and amplitude. Tsc1KO neurons were the only investigated condition that displayed increased sEPSC frequency. This finding corroborates a previous study showing increased sEPSC frequency with no sEPSC amplitude changes in layer 2/3 cortical pyramidal neurons from 4-week-old Tsc1 conditional KO mice58. Interestingly, as previously reported for RhebS16H neurons59, the sEPSC frequency in RhebY35L neurons was reduced by 25% but this did not reach statistical significance, possibly due to a generally weaker effect of the Y35L mutation compared to the S16H mutation. RhebY35L, mTORS2215Y, and PtenKO neurons exhibited increased sEPSC amplitudes. We found no changes in the sEPSC amplitude of Depdc5KOneurons, which differs from a previous study reporting increased sEPSC amplitude in these neurons26. This discrepancy may likely be explained by the differences in the age of the animals examined in their study (P20-24) versus ours (P26-43), given that dynamic changes in synaptic properties are still ongoing past P2160. Despite the above differences in sEPSC frequencies and amplitudes, the total charge was increased in almost all conditions except for in RhebY35L, suggesting enhanced synaptic drive and connectivity. These excitatory synaptic changes may counteract the increases in rheobase (i.e., decreased membrane excitability) and thereby impact circuit hyperexcitability and seizure vulnerability. Overall, the changes in synaptic activity for the RhebY35L condition are in stark contrast to the other gene conditions. These observations suggest a Rheb-specific impact on synaptic activity that differs from the other mTORC1 pathway genes. Thus, despite impinging on mTORC1 signaling, different mTORC1 pathway gene mutations can affect synaptic activity, and thereby, excitability differently. The mechanisms accounting for the observed differences are not clear, but it is increasingly acknowledged that additional pathways beyond mTORC1 are activated by each gene condition that could contribute to the differential impacts on synaptic activity13. One limitation of this study is that we did not examine inhibitory synaptic properties and the impact on excitatory-inhibitory balance. These changes could potentially contribute to gene-specific mechanisms of cortical hyperexcitability and should be the subject of another study.

In summary, we have shown that mutations affecting different mTORC1 pathway genes have similar and dissimilar consequences on cortical pyramidal neuron development and function, which may affect how neurons behave in cortical circuits. Our findings suggest that cortical neurons harboring different mTORC1 pathway gene mutations may differentially affect how neurons receive and process cortical inputs, which have implications for the mechanisms of cortical hyperexcitability and seizures in FMCDs, and potentially affect how neurons and their networks respond to therapeutic intervention.

Author contribution

LN: conceptualization, investigation, analysis, visualization, writing-original draft, review & editing, YX: investigation, analysis, MN: analysis, AB: conceptualization, writing-original draft, review & editing.

Data availability

The datasets generated during and/or analyzed during the current study are available from the corresponding authors upon reasonable request.

Competing interest

AB is an inventor on a patent application “Methods of treating and diagnosing epilepsy” (Pub. no. US2022/0143219 A1).

Grant funding

National Institutes of Health (NIH) F32 HD095567 (LHN), R01 NS111980 (AB)

eLife Assessment

This manuscript examines shared and divergent mechanisms of disruptions of five different mTOR pathway genes on embryonic mouse brain neuronal development. The significance of the manuscript is important, because it bridges several different genetic causes of focal malformations of cortical development. The strength of evidence is compelling, relying on both gain and loss of function, demonstrating differential impact on excitatory synaptic activity, conferring gene-specific mechanisms of hyperexcitability. The results have both theoretical and practical implications for the field of developmental neurobiology and clinical epilepsy.

Reviewer #1 (Public Review):

Pathogenic mutations of mTOR pathway genes have been identified in patients with malformation of cortical development and intractable epilepsy. Nguyen et al., established an in vivo rodent model to investigate the impact of different mTOR pathway gene dysfunction on neuronal intrinsic membrane excitability and cortical network activity. The results demonstrate that activation of mTORC1 activators or inactivation of mTORC1 repressors leads to convergent mTOR pathway activation and alterations of neuronal morphology, the key pathological feature of human FCD and hemimegalencephaly. However, different mTOR pathway gene mutations also exhibited variations in modulating Ih current and synaptic activity in rodent cortical neurons. These findings provide novel insights into the mechanism of seizure generation associated with cortical malformation.

1. The authors found differences in the initial spike doublet of action potentials between cortical neurons in experimental and control conditions (Figure 2e). The action potential firing frequency of the first two APs (instant firing frequency) of recorded neurons shall be quantified to investigate whether there are statistical differences between the action potential firing frequency in cortical neurons in different experimental groups versus control conditions.

2. The mTORS12215Y induced the largest changes in Ih current amplitudes in cortical neurons compared with other experimental conditions. Whether the HCN4 channel expression is regulated by mTOR pathway activation, or could there be possible interactions between the HCN channel and mTORS12215Y mutant protein?

3. A comparison of the electrophysiological characteristics of cortical neurons in different experimental conditions in the present study and pathological neurons in human FCD reported in previous literature could be interesting. Inducing pathological gene mutations or knocking out key genes in mTOR pathway in the rodent cortex - which approach could better model human FCD?

Reviewer #2 (Public Review):

Summary:

The study provides valuable and compelling evidence that while activation of the mTOR cascade confers some similarities in alterations in cell size, mTOR pathway activation, cortical lamination, baseline firing properties, and synaptic activity, there are distinctions that could account for clinical differences in seizure and epilepsy phenotypes in patients harboring these mutations. These findings could have important implications going forward as we design clinical therapeutic strategies to modulate mTOR activity in these individuals to treat seizures.

This study presents a valuable finding on the role that distinct mTOR pathway genes play in altered cell shape, cortical laminar migration, and cellular excitability in the mouse medial prefrontal cortex (mPFC). The evidence supporting the claims of the authors is solid, although analysis of the role of the mTORC2 pathway and consideration of distinct metabolic states i.e., amino acid levels would have strengthened the study. The work will be of interest to neuroscientists working on human epilepsy. These genes have each been assayed in previous independent studies and thus the direct comparison is what provides the innovation and interest.

The manuscript by Nguyen and colleagues attempts to define both the common and differential roles of mTOR pathway genes, both by gene knockout (KO) and activation, on cortical neuronal size, cortical lamination, and excitability. They focused on 5 genes that have been linked to human malformations of cortical development (MCD) and epilepsy: RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO. The RhebY35L, mTORS2215Y are known and pathogenic human gain-of-function variants. Each of these genes is known to modulate the activity of mTORC1 and either KO or activation will lead to abnormal and persistent hyperactivation of mTOR activity. Using in utero electroporation they transfected plasmids containing these gene constructs into fetal mouse brains at E15.5 and then assessed neuronal shape and size, laminar positioning, spontaneous activity, synaptic activity, and expression of a novel voltage-gated potassium channel (HCN4) at varying time postnatally e.g., P7-9 (neonates) and P28-43 (young adults).

The study clearly achieves its stated aims i.e., that disruption of each of five distinct mTOR pathway genes, Rheb, mTOR, Depdc5, Pten, and Tsc1, individually impacts pyramidal neuron development and electrophysiological function in the mouse mPFC. The data from each of the 5 genes provides strong support to the notion that mTOR pathway gene mutations yield the unifying clinical parcellation of mTORopathies, likely as a consequence of mTOR pathway activation. The data also provide interesting evidence that subtle or even overt differences in clinical phenotypes between RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO in humans could be due to effects of these genes either on mTOR or on yet to be defined alternative pathways. Assuredly follow-up studies to examine how Rheb, mTOR, Dedpdc5, Pten, and Tsc1 engage with other protein binding partners or other pathways will be warranted in future studies.

Strengths:

The investigators demonstrate that gene KO or activation leads to common changes in cell size (enlargement) though with different effects across each gene subtype suggesting distinct genetic effects despite a common effect on mTOR signaling. The major effect was seen in forebrain neurons expressing mTORS2215Y. They also report gene-specific effects of each mTOR pathway gene on cortical lamination. For example, while RhebY35L, mTORS2215Y, Dedpdc5KO, and Tsc1KO significantly disrupted laminar positioning of neurons in layer 2/3, PtenKO had minimal effects on laminar positioning. This finding is intriguing since it means that simply activating mTOR during fetal brain development will not necessarily alter cortical lamination and that an increase in cell size by itself doesn't disrupt laminar fidelity. To verify that the expression of plasmids led to mTORC1 hyperactivation, phosphorylated levels of S6 (i.e., p-S6), a downstream substrate of mTORC1, were assayed by immunohistochemistry in P28-43 mice. Expression of the RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO plasmids all led to significantly increased p-S6 staining intensity, supporting that the expression of each of these plasmids leads to increased mTORC1 signaling.

Whole-cell current- and voltage-clamp recordings were performed in P25-P51 mice in acute brain slice preparations. Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO led to decreased depolarization-induced excitability, but only RhebY35L, mTORS2215Y, and Tsc1KO expression led to depolarized resting membrane potentials. Interestingly, expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO led to the abnormal presence of HCN4 channels with variations in functional expression suggesting a common cellular mechanism that could confer excitability. Treatment with rapamycin, an mTOR inhibitor, reversed the expression changes in HCN4. Expression of RhebY35L, mTORS2215Y, Dedpdc5KO, PtenKO, and Tsc1KO led to different impacts on sEPSC properties. Effects of treatment with the selective HCN channel blocker zatebradine on hyperpolarization-induced inward currents in mTORS2215Y neurons confirmed the identity of ΔI as Ih.

Overall the data presented provides a convincing and compelling direct comparison of the roles that select mTOR pathway genes play on brain development and network excitability. It is critical to directly compare these gene effects in mouse models because although these genes are part of the mTOR pathway and clearly cause augmentation of mTOR activation, there are mechanistic differences in how these gees modify mTOR and how they interact with other proteins and phenotypic differences in humans harboring mutations in these same genes.

Weaknesses:

There are a few limitations to an otherwise solid study. First, the authors postulate that all the findings are dependent on mTORC1-related effects but don't assess whether some of the differences could be due to effects on mTORC2 signaling. mTORC2 is an important and poorly understood alternative isoform of mTOR (due to rictor binding) that has effects on distinct cell signaling pathways and in particular actin polymerization. This doesn't diminish the effects of the current analysis of mTORC1 but could explain genotypic differences in each variable. A few prior studies have assessed the role of mTORC2 in epileptogenesis and Cortical malformations (Chen et al., 2019)

Second, the slice recordings were performed in the usual recording aCSF buffer conditions but there is no assessment of the role of amino acids or nutrients in the bath. While it is clear that valuable and viable acute slice recordings can be made in aCSF, the role of the mTOR pathway is to modulate cell growth in response to nutrient conditions. Thus, one variable that could be manipulated and assessed currently in this study is the levels of amino acids i.e., leucine and arginine added to the bath since DEPDC5 and TSC1 are responsive to ambient amino acid levels.

Third, the analysis concedes that the role of somatic mutations in cortical malformations may depend not only on genotypic effects but also on allelic load and cellular subtype affected by the mutation. Thus, it would interesting to see if electroporation either at E14 or E16, thereby affecting a distinct pool of progenitors, would mitigate or accentuate differences between mTOR pathway genes.

Treatment with rapamycin and zatebradine in each condition would have added to the strength of the findings to determine the mTOR-dependence and reversibility of HCN4 effects.

Author Response

The following is the authors’ response to the original reviews.

Public Reviews:

We thank the reviewers for their insightful comments on our manuscript. We have addressed the reviewers’ comments below and in the revised manuscript.

Reviewer #1:

Comment #1: The authors found differences in the initial spike doublet of action potentials between cortical neurons in experimental and control conditions (Figure 2e). The action potential firing frequency of the first two APs (instant firing frequency) of recorded neurons shall be quantified to investigate whether there are statistical differences between the action potential firing frequency in cortical neurons in different experimental groups versus control conditions.

Response: As suggested by the reviewer, we have quantified the first interspike interval (ISI; time between the 1st and 2nd action potential). The data is included in Fig. 2h as well as in Fig. S3e and Table 1. The Results and Methods have also been updated accordingly.

Comment #2: The mTORS12215Y induced the largest changes in Ih current amplitudes in cortical neurons compared with other experimental conditions. Whether the HCN4 channel expression is regulated by mTOR pathway activation, or could there be possible interactions between the HCN channel and mTORS12215Y mutant protein?

Response: Our previous findings using the RhebS16H mutation support the idea that increased expression of HCN4 channels is regulated by mTOR pathway activation. This is evidenced by its sensitivity to rapamycin (a mTOR inhibitor) and expression of constitutively active 4E-BP1 (a translational repressor downstream of mTORC1). Since mTORS2215Y directly hyperactivates mTORC1 and there are no known interactions between HCN channels and mTORS2215Y, our data strongly suggests that abnormal HCN4 channel expression occurs via mTORC1 hyperactivation in this condition. We have revised our Discussion to make this point clearer.

Comment #3: A comparison of the electrophysiological characteristics of cortical neurons in different experimental conditions in the present study and pathological neurons in human FCD reported in previous literature could be interesting. Inducing pathological gene mutations or knocking out key genes in mTOR pathway in the rodent cortex - which approach could better model human FCD?

Response: We agree with the reviewer and have added a new paragraph in the Discussion to compare our electrophysiology results to those of previous studies done on human FCDII and TSC cytomegalic neurons. With regards to the reviewer’s question about which of the two approaches in the rodent cortex – inducing pathological gene mutations or knocking out key genes in the mTOR pathway – would better model human FCD, our study emphasizes the importance of considering gene-specific mechanisms in FCDII. Thus, modeling the genetically distinct FCDIIs will require using gene-specific manipulations. We have revised our Discussion to include this point. With that said, for some phenotypes that are generalized across FCDII independent of the mTOR pathway genes, using pathogenic mutations of mTOR activators or knockout of negative mTOR regulators would likely both be appropriate models. Of note, as discussed in the manuscript, there are also technical factors to be considered when choosing to use a pathogenic gene mutation versus knocking out a gene (the latter which would depend on the half-life of the proteins).

Reviewer #2:

Comment #1: The authors postulate that all the findings are dependent on mTORC1-related effects but don't assess whether some of the differences could be due to effects on mTORC2 signaling. mTORC2 is an important and poorly understood alternative isoform of mTOR (due to rictor binding) that has effects on distinct cell signaling pathways and in particular actin polymerization. This doesn't diminish the effects of the current analysis of mTORC1 but could explain genotypic differences in each variable. A few prior studies have assessed the role of mTORC2 in epileptogenesis and cortical malformations (Chen et al., 2019).

Response: We agree with the reviewer and have revised our Discussion to include the possibility of mTORC2 contribution to the gene-specific phenotypic differences.

Comment #2: The slice recordings were performed in the usual recording aCSF buffer conditions but there is no assessment of the role of amino acids or nutrients in the bath. While it is clear that valuable and viable acute slice recordings can be made in aCSF, the role of the mTOR pathway is to modulate cell growth in response to nutrient conditions. Thus, one variable that could be manipulated and assessed currently in this study is the levels of amino acids i.e., leucine and arginine added to the bath since DEPDC5 and TSC1 are responsive to ambient amino acid levels.

Response: We thank the reviewer for this great suggestion, and we intend to pursue this as part of another study.

Comment #3: The analysis concedes that the role of somatic mutations in cortical malformations may depend not only on genotypic effects but also on allelic load and cellular subtype affected by the mutation. Thus, it would be interesting to see if electroporation either at E14 or E16, thereby affecting a distinct pool of progenitors, would mitigate or accentuate differences between mTOR pathway genes.

Response: We agree with the reviewer. This is a crucial experiment that we hope to perform in the future. We have also added a paragraph in our Discussion to address this important point.

Comment #4: Treatment with rapamycin and zatebradine in each condition would have added to the strength of the findings to determine the mTOR-dependence and reversibility of HCN4 effects.

Response: We previously demonstrated the mTORC1 dependence of HCN4 expression in the RhebS16H condition using rapamycin and expression of constitutively active 4E-BP1. 4E-BP1 is a translational repressor downstream of mTORC1. In the 4E-BP1 study, we used a conditional system to express 4EBP1F113A (mutation that resists inactivation by mTORC1) in adolescent mice while RhebS16H (and thus mTORC1 activation) was expressed embryonically. 4E-BP1F113A expression suppressed Ih current and HCN4 expression, suggesting that aberrant HCN4 expression can be reversed by decreasing mTORC1regulated translation. Based on these data and the findings that rapamycin suppressed abnormal HCN4 expression, we postulate that increased HCN4 expression in the different gene conditions examined in the present study occurs via the mTORC1 pathway. However, we agree with the reviewer that treating each of the conditions with rapamycin would provide direct evidence of their mTORC1 dependence. Additionally, treating each condition with the HCN channel blocker zatebradine would also add strength to the findings. We have added a comment in the Discussion to acknowledge this point.

Reviewer #1 (Recommendations For The Authors):

Comment #1: The authors found increased frequency or amplitudes of spontaneous postsynaptic currents in different experimental cohorts. These data may not be sufficient to conclude increased synaptic excitability, because there are no pharmacological experiments to verify whether the recorded inward currents are excitatory or inhibitory postsynaptic currents. An alternative approach could be analyzing the decay time of spontaneous postsynaptic currents, the excitatory postsynaptic currents had relatively faster decay time compared with inhibitory postsynaptic currents.

Response: Thank you for the comment. We apologize for the lack of clarity and have added the following text in the Results to clarify: “To separate sEPSCs from spontaneous inhibitory postsynaptic currents (sIPSCs), we used an intracellular solution rich in K-gluconate to impose a low intracellular Cl- concentration and recorded at a holding potential of -70 mV, which is near the Cl- reversal potential. The 90%-10% decay time of the measured synaptic currents ranged between 4-8 ms in all conditions (mean ± SD: control: 4.9 ± 1.6; RhebY35L: 5.2 ± 1.4; mTORS2215Y: 7.4 ± 1.4; control: 6.8 ± 0.7; Depdc5KO: 7.4 ± 1.0; PtenKO: 8.1 ± 0.9; Tsc1KO: 7.4 ± 0.9), consistent with the expected decay time for sEPSCs and shorter than the decay time for sIPSCs (Kroon et al, 2019). The recorded synaptic currents were therefore considered to be sEPSCs.”

Comment #2: There are typos of Depdc5 in the text and figure legends.

Response: Thank you for noticing this error. We have corrected the typos in the manuscript.

Version 2

Abstract

Abstract

Brain somatic mutations in various components of the mTOR complex 1 (mTORC1) pathway have emerged as major causes of focal malformations of cortical development and intractable epilepsy. While these distinct gene mutations converge on excessive mTORC1 signaling and lead to common clinical manifestations, it remains unclear whether they cause similar cellular and synaptic disruptions underlying cortical network hyperexcitability. Here, we show that in utero activation of the mTORC1 activators, Rheb or mTOR, or biallelic inactivation of the mTORC1 repressors, Depdc5, Tsc1, or Pten in mouse medial prefrontal cortex leads to shared alterations in pyramidal neuron morphology, positioning, and membrane excitability but different changes in excitatory synaptic transmission. Our findings suggest that, despite converging on mTORC1 signaling, mutations in different mTORC1 pathway genes differentially impact cortical excitatory synaptic activity, which may confer gene-specific mechanisms of hyperexcitability and responses to therapeutic intervention.

Introduction

Focal malformations of cortical development (FMCDs), including focal cortical dysplasia type II (FCDII) and hemimegalencephaly (HME), are the most common causes of intractable epilepsy in children1,2. These disorders are characterized by abnormal brain cytoarchitecture with cortical overgrowth, mislamination, and cellular anomalies, and are often associated with developmental delay and intellectual disability3. Early immunohistochemical studies identified hyperactivation of the mechanistic target of rapamycin complex 1 (mTORC1) signaling pathway in resected brain tissue from individuals with FCDII and HME, leading to the classification of these disorders as “mTORopathies”4. More recently, somatic mutations in numerous regulators of the mTORC1 pathway were identified as the genetic causes of FCDII and HME5–7. Accumulating evidence shows that these mutations occur in a subset of dorsal telencephalic progenitor cells that give rise to excitatory neurons during fetal development, resulting in brain somatic mosaicism8,9. These genetic findings provide opportunities for gene therapy approaches targeting the mTORC1 pathway for FCDII and HME.

mTORC1 is an evolutionarily conserved protein complex that promotes cell growth and differentiation through the regulation of protein synthesis, metabolism, and autophagy10. mTORC1 is composed of mTOR, a serine/threonine kinase that exerts the complex’s catalytic function, Raptor, PRAS40, and mLST8. Activation of mTORC1 is controlled by a well-described upstream cascade involving multiple protein regulators11. Stimulation by growth factors activates phosphoinositide 3-kinase (PI3K), which leads to the activation of phosphoinositide-dependent kinase 1 (PDK1) and subsequent phosphorylation and activation of AKT. Activated AKT phosphorylates and inactivates tuberous sclerosis complex 1/2 (TSC1/2 complex; consisting of TSC1 and TSC2 proteins), which releases the brake on Ras homolog enriched in brain (RHEB), a GTP-binding protein that directly activates mTORC1. This mTORC1-activating cascade is negatively controlled by the phosphatase and tensin homolog (PTEN) protein, which inhibits PI3K activation of PDK1. Additionally, mTORC1 signaling is regulated by a separate nutrient-sensing GAP Activity Towards Rags 1 (GATOR1) complex, which consists of DEP domain containing 5 (DEPDC5) and the nitrogen permease regulator 2-like (NPRL2) and 3-like (NPRL3) proteins12. The GATOR1 complex serves as a negative regulator of mTORC1 that inhibits mTORC1 at low amino acid levels.

Pathogenic mutations in numerous regulators that activate mTORC1 signaling, including PIK3CA, PTEN, AKT3, TSC1, TSC2, MTOR, RHEB, DEPDC5, NPRL2, and NPRL3, have been identified in HME and FCDII9. Genetic targeting of these genes in mouse models consistently recapitulates the epilepsy phenotype, supporting an important role for these genes in seizure generation13,14. However, while all these genes impinge on the mTORC1 pathway, many of them also participate in mTORC1-independent functions through parallel signaling pathways13,15–19, and it is unclear whether mutations affecting different mTORC1 pathway genes lead to cortical hyperexcitability through common neural mechanisms. In this study, we examined how disruption of five distinct mTORC1 pathway genes, Rheb, mTOR, Depdc5, Pten, and Tsc1, individually impact pyramidal neuron development and electrophysiological function in the mouse medial prefrontal cortex (mPFC). Collectively, we found that activation of the mTORC1 activators, Rheb and mTOR, or inactivation of the mTORC1 repressors, Depdc5, Tsc1, and Pten, largely leads to similar alterations in neuron morphology and membrane excitability but differentially impacts excitatory synaptic activity. The latter has implications for cortical network function and seizure vulnerability and may affect how individuals with different genotypes respond to targeted therapeutics.

Results

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to varying magnitudes of neuronal enlargement and mispositioning in the cortex

To model somatic gain-of-function mutations in the mTORC1 activators, we individually expressed plasmids encoding RhebY35L or mTORS2215Y, respectively, in select mouse neural progenitor cells during late corticogenesis, at embryonic day (E) 15, via in utero electroporation (IUE) (Fig. 1a, b). These pathogenic variants were previously detected in brain tissue from children with FCDII and HME associated with intractable seizures20–25. We specifically targeted a subset of late-born progenitor cells that generate excitatory pyramidal neurons destined to layer 2/3 in the medial prefrontal cortex (mPFC) to mimic frontal lobe somatic mutations and the genetic mosaicism characteristic of these lesions. To model the somatic loss-of-function mutations in mTORC1 repressors, we expressed plasmids encoding previously validated CRISPR/Cas9 guide RNAs against Depdc526, Tsc127, or Pten28 to individually knockout (KO) the respective genes using the same IUE approach (Fig. 1a, b). As a control for the activating mutations, we expressed plasmids encoding fluorescent proteins under the same CAG promoter. As a control for the CRISPR/Cas9-mediated knockouts, we used an empty CRISPR/Cas9 vector containing no guide RNA sequences. To verify that the expression of these plasmids leads to mTORC1 hyperactivation, we assessed the phosphorylated levels of S6 (i.e., p-S6), a downstream substrate of mTORC1, using immunostaining in brain sections from postnatal day (P) 28-43 mice. As expected, we found that expression of the RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO plasmids all led to significantly increased p-S6 staining intensity, supporting that the expression of each of these plasmids leads to increased mTORC1 signaling (Fig. 1c, d, Table 1, Fig. S1a).

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to varying magnitudes of neuronal enlargement and mispositioning in the cortex.

(a) Diagram of the PI3K-mTORC1 pathway. Activation of mTORC1 signaling is controlled by positive (blue) and negative (purple) regulators within the pathway. (b) Diagram of overall experimental timeline and approach. IUE was performed at E15.5. A cohort of animals was used for patch clamp recording at P26-51 and another cohort was used for histology at P28-43. (c) Representative images of tdTomato+ cells (red) and p-S6 staining (green) in mouse mPFC at P28-43. (d) Quantification of p-S6 staining intensity (normalized to the mean control) in tdTomato+ neurons. (e) Quantification of tdTomato+ neuron soma size. (f) Representative images of tdTomato+ neuron (red) placement and distribution in coronal mPFC sections. Red square on the diagram denotes the imaged area for all groups. CC, corpus callosum. (g) Quantification of tdTomato+ neuron placement in layer 2/3. Left diagram depicts the approach for analysis: the total number of tdTomato+ neurons within layer 2/3 (white square) was counted and expressed as a % of total neurons in the imaged area. Right bar graphs show the quantification. (h) Quantification of tdTomato+ neuron distribution across cortical layers. Left diagram depicts the approach for analysis: the imaged area was divided into 10 equal-sized bins across the cortex, spanning the corpus callosum to the pial surface (white grids); the total number of tdTomato+ neurons within each bin was counted and expressed as a % of total neurons in the imaged area. Right bar graphs show the quantification. For graphs d, e: n = 4-8 mice per group, with 6-15 cells analyzed per animal. For graphs g, h: n = 3-7 mice per group, with 1 brain section analyzed per animal. Statistical comparisons were performed using (d, e) nested one-way ANOVA (fitted to a mixed-effects model) to account for correlated data within individual animals, (g) one-way ANOVA, or (h) two-way repeated measures ANOVA. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001. All data are reported as the mean of all neurons or brain sections in each group ± SEM.

Considering the cytoarchitectural abnormalities associated with mTORC1 hyperactivation, we compared neuron soma size and positioning in the cortex following in utero activation of Rheb and mTOR and inactivation of Depdc5, Pten, and Tsc1 in P28-43 mice. While all experimental conditions led to increased neuron soma size, the magnitude of the enlargement was dependent on the specific gene that was targeted (Fig. 1c, e, Table 1. Fig. S1b). In particular, expression of RhebY35L and mTORS2215Y led to the largest soma size changes, with a >3-fold increase from control. Expression of PtenKOled to a similarly large increase of 2.7-fold, while expression of Depdc5KOand Tsc1KO led to a 1.7 and 2.1-fold increase, respectively. The increase in the PtenKO condition was significantly higher than both the Depdc5KO and Tsc1KO conditions, and the increase in the Tsc1KO condition was significantly higher than the Depdc5KO condition. Although the above analysis was performed at P28-43, the enlargement of neuron soma sizes was already detected by P7-9 (Fig. S2a-c, Table S1). In terms of neuronal positioning, all experimental conditions, except for PtenKO, resulted in neuron misplacement (Fig. 1f-h, Table 1). The RhebY35L and mTORS2215Yconditions led to the most severe phenotype: ∼70-80% of the neurons were misplaced outside of layer 2/3 (Fig. 1g), with the mispositioned neurons evenly scattered across the deeper layers (Fig. 1h). For the Depdc5KO and Tsc1KO conditions, ∼45-60% of the neurons were misplaced outside of layer 2/3 (Fig. 1g), with the mispositioned neurons found closer to layer 2/3 (Fig. 1h). Taken together, these studies show that the expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to varying magnitudes of neuronal enlargement and mispositioning in the cortex. Of note, PtenKO neurons, despite exhibiting a 2.7-fold increase in neuron soma size, were mostly correctly positioned in layer 2/3. These findings suggest that while all experimental conditions lead to increased soma size, not all lead to neuron mispositioning, suggesting defective migration and the subsequent impact on neuron positioning occur independently of cell size.

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO universally leads to decreased depolarization-induced excitability, but only RhebY35L, mTORS2215Y, and Tsc1KO expression leads to depolarized resting membrane potential

To elucidate the contribution of each experimental condition to the function of cortical neurons, we obtained whole-cell patch clamp recordings of layer 2/3 pyramidal neurons at P26-P51 (Fig. 2a). The RhebY35Land mTORS2215Yconditions were compared to a control group expressing fluorescent proteins under the same CAG promoter. The Depdc5KO, PtenKO, and Tsc1KO conditions were compared to a CRISPR/Cas9 empty vector control. Recordings of control and experimental conditions were alternated to match the animal ages.

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO universally leads to decreased depolarization-induced excitability, but only RhebY35L, mTORS2215Y, and Tsc1KO expression leads to depolarized RMPs.

(a) Diagram of experimental approach for whole-cell patch clamp recording. Recordings were performed in layer 2/3 pyramidal neurons in acute coronal slices from P26-51 mice, expressing either control, RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, or Tsc1KO plasmids. (b-d) Graphs of membrane capacitance, resting membrane conductance, and RMP. (e) Representative traces of the AP firing response to a 300 pA depolarizing current injection. (f) Input-output curves showing the mean number of APs fired in response to 500 ms-long depolarizing current steps from 0 to 400 pA. Arrows point to initial spike doublets. (g-k) Graphs of rheobase, 1st ISI, AP threshold, AP peak amplitude, and AP half-width. For all graphs: n = 5-10 mice per group, with 16-50 cells analyzed per animal. Statistical comparisons were performed using (b-d, g-k) nested one-way ANOVA (fitted to a mixed-effects model) to account for correlated data within individual animals or (f) mixed-effects ANOVA accounting for repeated measures. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001, for the input-output curves in f: *p<0.05 (vs. control), #<0.05 (vs. Depdc5KO), Ɏp<0.05 (vs. Tsc1KO). All data are reported as the mean of all neurons in each group ± SEM.

Consistent with the changes in soma size (Fig. 1c, e), recorded neurons displayed increased membrane capacitance in all experimental conditions (Fig. 2b, Table 1, Fig. S3a). Neurons expressing the mTORS2215Y variant had a larger capacitance increase than those expressing the RhebY35L variant. PtenKO and Tsc1KO neurons had similar increases in capacitance that were both larger than that of the Depdc5KO neurons. All neurons across the experimental conditions also had increased resting membrane conductance in a pattern that followed that of the capacitance (Fig. 2c, Table 1, Fig. S3b). However, while RhebY35L, mTORS2215Y, and Tsc1KOexpression led to depolarized resting membrane potential (RMP), Depdc5KOand PtenKO expression did not significantly affect the RMP (Fig. 2d, Table 1, Fig. S3c). To assess whether these changes impacted neuron intrinsic excitability, we examined the action potential (AP) firing response to depolarizing current injections. For all experimental conditions, neurons fired fewer APs for current injections above 100 pA compared to the respective control neurons (Fig. 2e, f, Table 1). This decrease in intrinsic excitability is reflected in the increased rheobase (i.e., the minimum current required to induce an AP) in all experimental conditions, with the mTORS2215Y and RhebY35Lconditions leading to the largest rheobase increases (Fig. 2g, Table 1, Fig. S3d). Collectively, these findings indicate that RhebY35L, mTORS2215Y, and Tsc1KO neurons display a decreased ability to generate APs upon depolarization despite having depolarized RMPs. In terms of firing pattern, neurons in all groups displayed a regular-spiking pattern with spike-frequency adaptation (Fig. 2e). However, while an initial spike doublet was observed in control neurons, consistent with the expected firing pattern for layer 2/3 mPFC pyramidal neurons29, this was absent in all the experimental conditions except for the PtenKO condition (Fig. 2e). Further quantification of the first interspike interval (1st ISI; interval between the 1st and 2nd AP) revealed significantly increased 1st ISI in RhebY35L, mTORS2215Y, Depdc5KO, and Tsc1KO neurons, but not in PtenKO neurons, compared to control neurons (Fig. 2h, Table 1, Fig. S3e). No differences in the AP threshold were observed across the conditions (Fig. 2i, Table 1, Fig. S3f). The AP peak amplitude was decreased in the mTORS2215Y and Tsc1KO conditions (Fig. 2j, Table 1, Fig. S3g), while the AP half-width was decreased in the RhebY35L, mTORS2215Y, and PtenKO conditions (Fig. 2k, Table 1, Fig. S3h). Taken together, these findings show that various genetic conditions that activate the mTORC1 pathway universally lead to decreased depolarization-induced excitability in layer 2/3 pyramidal neurons, with gene-dependent changes in RMP and several AP properties.

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to the abnormal presence of HCN4 channels with variations in functional expression

We recently reported that neurons expressing RhebS16H, an mTOR-activating variant of Rheb, display abnormal expression of HCN4 channels30,31. These channels give rise to a hyperpolarization-activated cation current (Ih) that is normally absent in layer 2/3 pyramidal neurons30,31. The aberrant Ih, which has implications for neuronal excitability, preceded seizure onset and was detected by P8-12 in mice30. Rapamycin treatment starting at P1 and expression of constitutive active 4E-BP1, a translational repressor downstream of mTORC1, prevented and alleviated the aberrant HCN4 channel expression, respectively30,31. These findings suggest that the abnormal HCN4 channel expression is mTORC1-dependent. Given that all the experimental conditions in the present study led to increased mTORC1 activity, we investigated whether they also resulted in abnormal HCN4 channel expression. Immunostaining for HCN4 channels using previously validated antibodies30 in brain sections from P28-43 mice revealed significantly increased HCN4 staining intensity in the electroporated neurons in all experimental conditions compared to the respective controls, which exhibited no HCN4 staining (Fig 3a, b, Table 1, Fig. S4a). The increased staining was evident in the ipsilateral cortex containing mTORS2215Yelectroporated neurons and absent in the non-electroporated contralateral cortex (Fig. S5a). These data indicate the presence of aberrant HCN4 channel expression following RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, or Tsc1KO expression in layer 2/3 pyramidal neurons.

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to the abnormal presence of HCN4 channels with variations in functional expression.

(a) Representative images of tdTomato+ cells (red) and HCN4 staining (green) in mouse mPFC at P28-43. (b) Quantification of HCN4 intensity (normalized to the mean control) in tdTomato+ neurons. (c) Representative current traces in response to a series of 3 s-long hyperpolarizing voltage steps from -120 to -40 mV, with a holding potential of -70 mV. Current traces from the -40 and -50 mV steps were not included due to contamination from unclamped Na+ spikes. Orange lines denote the current traces at -90 mV. (d) IV curve obtained from Iss amplitudes. (e) ΔIV curve obtained from Ih amplitudes (i.e., ΔI, where ΔI=Iss – Iinst). (f) Graphs of Ih amplitudes at –90 mV. (g) Representative current traces in response to a series of 3-s long hyperpolarizing voltage steps from -110 mV to -50 mV in the mTORS2215Y condition pre- and post-zatebradine application. Orange lines denote the current traces at -90 mV. (h) IV curve obtained from Iss amplitudes in the mTORS2215Y condition pre- and post-zatebradine application. (i) ΔIV curve obtained from Ih amplitudes (i.e., ΔI) in the mTORS2215Y condition pre- and post-zatebradine application. Arrow points to the post-zatebradine Ih amplitude at -90 mV. (j) Representative traces of the zatebradine-sensitive current obtained after subtraction of the post-from the pre-zatebradine current traces in response to -110 mV to -50 mV voltage steps. Orange lines denote the current traces at -90 mV. (k) IV curve of the zatebradine-sensitive current obtained after subtraction of the post-from the pre-zatebradine current traces. (l) Graph of RMP in the control and mTORS2215Y conditions pre- and post-zatebradine application. Connecting lines denote paired values from the same neuron. For graph b: n = 4-8 mice per group, with 4-15 cells analyzed per animal. For graphs d, e, f: n = 5-10 mice per group, with 24-47 cells analyzed per animal. For graphs h, i, k, l: n = 4-6 neurons (paired). Statistical comparisons were performed using (b, f) nested ANOVA (fitted to a mixed-effects model) to account for correlated data within individual animals, (d, e) mixed-effects ANOVA accounting for repeated measures, or (h, i, l) two-way repeated measures ANOVA. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ****p<0.0001, for the IV curves in d, e, h, i: *p<0.05 (vs. control), #p<0.05 (vs. RhebY35Lor Depdc5KO), Ɏp<0.05 (vs. PtenKO). All data are reported as the mean of all neurons in each group ± SEM.

To examine the functional impacts of the aberrant HCN4 channel expression, we examined Ih amplitudes in the various experimental conditions. To evoke Ih, we applied a series of 3 s-long hyperpolarizing voltage steps from -120 mV to -40 mV. Consistent with previous findings in RhebS16H neurons30,31, hyperpolarizing voltage pulses elicited significantly larger inward currents in all experimental conditions compared to their respective controls (Fig. 3c, d, Table 1). The mTORS2215Ycondition displayed larger inward currents than the RhebY35L condition, while the PtenKO condition displayed the largest inward currents compared to the Depdc5KO and Tsc1KO conditions (Fig. 3d). These data were proportional to the changes in neuron soma size (Fig. 1e, 2b). The inward currents in RhebS16H neurons are thought to result from the activation of both inward-rectifier K+ (Kir) channels and HCN channels30. Kir-mediated currents activate fast whereas HCN-mediated currents, i.e., Ih, activate slowly during hyperpolarizing steps; therefore, to assess Ih amplitudes, we measured the difference between the instantaneous and steady-state currents at the beginning and end of the voltage pulses, respectively (i.e., ΔI)32. The resulting ΔI-voltage (V) curve revealed significantly larger Ih amplitudes in all experimental conditions (Fig. 3e, Table 1). To further isolate the Ih from Kir -mediated currents, we measured Ih amplitudes at - 90 mV, near the reversal potential of Kir channels to eliminate Kir-mediated current contamination. At -90 mV, Ih amplitudes were significantly higher in the mTORS2215Y and Tsc1KO conditions compared to controls (Fig. 3f, Table 1, Fig. S4b). Of note, although the Depdc5KO and PtenKO conditions did not display a significant increase in Ih amplitudes at -90 mV, 1 out of 28 Depdc5KO neurons and 4 out of 27 PtenKO neurons had Ih amplitudes that were 2-fold greater than the mean Ih amplitude of the Tsc1KO condition. 6 out of 24 RhebY35L neurons also had Ih amplitudes 2-fold greater than this value (Fig. 3f). These data suggest that Ih currents are present in a subset of RhebY35L, Depdc5KO, and PtenKO neurons, and most mTORS2215Y and Tsc1KO neurons.

Considering that the mTORS2215Y condition led to the largest Ih, we examined the impact of the selective HCN channel blocker zatebradine on hyperpolarization-induced inward currents in mTORS2215Y neurons to further confirm the identity of ΔI as Ih. Application of 40 µM zatebradine reduced the overall inward currents (Fig. 3g, h, Table 1) and ΔI (Fig. 3i, Table 1). At -90 mV, ΔI was significantly decreased from -167.7 ± 54.2 pA to 0.75 ± 8.2 pA (± SD) (Fig. 3i, arrow). Subtraction of the post-from the pre-zatebradine current traces isolated zatebradine-sensitive inward currents which reversed near -50 mV, as previously reported for HCN4 channel reversal potentials33 (Fig. 3j, k, Table 1). These experiments verified the identity of ΔI as Ih. In comparison, application of the Kir channel blocker barium chloride (BaCl2) substantially reduced the overall inward currents but had no effects on ΔI (i.e., Ih) in Tsc1KOneurons (Fig. S6a-i). Consistent with the function of Ih in maintaining RMP at depolarized levels34–38, application of zatebradine hyperpolarized RMP in mTORS2215Y neurons but did not affect the RMP of control neurons that exhibited no Ih (Fig. 3l, Table 1). Collectively, these findings suggest that RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO expression in layer 2/3 pyramidal neurons lead to the abnormal presence of HCN4 channels with variations in functional expression.

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to different impacts on excitatory synaptic activity

As part of examining neuron excitability, we recorded spontaneous excitatory postsynaptic currents (sEPSCs) in all the gene conditions. To separate sEPSCs from spontaneous inhibitory postsynaptic currents (sIPSCs), we used an intracellular solution rich in K-gluconate to impose a low intracellular Cl- concentration and recorded at a holding potential of -70 mV, which is near the Cl- reversal potential. The 90%-10% decay time of the measured synaptic currents ranged between 4-8 ms in all conditions (mean ± SD: control: 4.9 ± 1.6; RhebY35L: 5.2 ± 1.4; mTORS2215Y: 7.4 ± 1.4; control: 6.8 ± 0.7; Depdc5KO: 7.4 ± 1.0; PtenKO: 8.1 ± 0.9; Tsc1KO: 7.4 ± 0.9), consistent with the expected decay time for sEPSCs and shorter than the decay time for sIPSCs29. The recorded synaptic currents were therefore considered to be sEPSCs. The sEPSCs frequency was unchanged in all experimental conditions except for the Tsc1KOcondition, where the sEPSCs frequency was significantly increased (Fig. 4a, b, Table 1, Fig. S7a). Unlike the other experimental conditions, the RhebY35Lcondition displayed a slight decrease in sEPSC frequency, consistent with previous findings in RhebS16H neurons; however, this did not reach statistical significance (Fig. 4b). The sEPSC amplitude was larger in the RhebY35L, mTORS2215Y, and PtenKOconditions (Fig. 4a, c, Table 1, Fig. S7b). Although the amplitudes were slightly larger in the Depdc5KO and Tsc1KO conditions, these changes were not significant (Fig. 4c). Thus, Tsc1KOneurons display increased sEPSC frequency but unchanged amplitude, while RhebY35L, mTORS2215Y,and PtenKO neurons display increased sEPSC amplitude but unchanged frequency. Finally, there was an increase in the sEPSC total charge in all experimental conditions except for the RhebY35L condition (Fig. 4d, Table 1, Fig. S7c). Collectively, these findings suggest all experimental conditions, except for RhebY35L, lead to increased synaptic excitability, with variable impact on sEPSC frequency and amplitude.

Expression of RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO leads to different impacts on sEPSC properties.

(a) Representative sEPSC traces recorded at a holding voltage of -70 mV. Top and bottom traces are from the same neuron. (b-d) Graphs of sEPSC frequency, amplitude, and total charge. For all graphs: n = 5-9 mice per group, with 17-34 cells analyzed per animal. Statistical comparisons were performed using a nested ANOVA (fitted to a mixed effects model) to account for correlated data within individual animals. Post-hoc analyses were performed using Holm-Šídák multiple comparison test. *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001. All data are reported as the mean of all neurons in each group ± SEM.

Discussion

In this unprecedented comparison study, we examined the impacts of several distinct epilepsy-associated mTORC1 pathway gene mutations on cortical pyramidal neuron development and their electrophysiological properties and synaptic integration in the mouse mPFC. Through a combination of IUE to model the genetic mosaicism of FMCDs, histological analyses, and patch-clamp electrophysiological recordings, we found that activation of either Rheb or mTOR or biallelic inactivation Depdc5, Tsc1, or Pten, all of which increase mTORC1 activity, largely leads to similar alterations in neuron morphology and membrane excitability but differentially impacts excitatory synaptic transmission. These findings have significant implications for understanding gene-specific mechanisms leading to cortical hyperexcitability and seizures in mTORC1-driven FMCDs and highlight the utility of personalized medicine dictated by patient gene variants. Furthermore, our study emphasizes the importance of considering gene-specific approaches when modeling genetically distinct mTORopathies for basic research studies.

Several histological phenotypes associated with mTORC1 hyperactivation were anticipated and confirmed in our studies. We found increased neuron soma size across all gene conditions, consistent with previous reports24,26,27,39–42. Additionally, all conditions, except for PtenKO, resulted in neuronal mispositioning in the mPFC, with RhebY35L and mTORS2215Y conditions being the most severe. Interestingly, PtenKO neurons were correctly placed in layer 2/3, despite having one of the largest soma size increases, suggesting that cell positioning is independent of cell size. The lack of mispositioning of PtenKO neurons was surprising as it is thought that increased mTORC1 activity leads to neuronal misplacement, and aberrant migration of PtenKO neurons has been reported in the hippocampus43. Given that PTEN has a long half-life, with a reported range of 5 to 20 hours or more depending on the cell type44–51, it is possible that following knockout at E15, the decreases in existing protein levels lagged behind the time window to affect neuronal migration. However, TSC1 also has a long half-life of ∼22 hours52,53, and knockout of Tsc1 at the same time point was sufficient to affect neuronal positioning. Mispositioning of neurons following Pten knockout in E15 rats was reported by one study, but this was a very mild phenotype compared to findings in the other gene conditions39. Therefore, these data suggest that PTEN has gene-specific biological mechanisms that contribute to the lack of severe neuronal mispositioning in the developing cortex. mTORS2215Yexhibited the strongest phenotypes in terms of neuron size and positioning, which is perhaps not surprising as the mTOR protein itself forms the catalytic subunit of mTORC1. Depdc5KO exhibited the mildest changes in these parameters. Unlike PTEN, TSC1, RHEB, and mTOR, which regulate mTORC1 via the canonical PI3K-mTORC1 pathway in response to growth factor stimulation, DEPDC5 regulates mTORC1 via the GATOR1 complex in response to changes in cellular amino acid levels54. The different modes of mTORC1 regulation by DEPDC5 may contribute to the differences in the severity of the phenotypes.

To assess the impact of RHEB, mTOR, DEPDC5, PTEN, and TSC1 disruption on cortical neuron excitability, we examined the intrinsic biophysical and synaptic properties of layer 2/3 pyramidal neurons in the mPFC. We found that all gene manipulations led to increased membrane conductance but decreased AP firing upon membrane depolarization, consistent with previous reports in the somatosensory cortex26,39,42,55–58 and in the mPFC for the RhebS16H condition30,31. Neurons in all experimental conditions required a larger depolarization to generate an AP (increased rheobase), likely due to their enlarged cell size. These findings seem to contradict the theory that seizure initiation originates from these neurons. However, inhibiting firing in cortical pyramidal neurons expressing the RhebS16H mutation via co-expression of Kir2.1 channels (to hyperpolarize neurons and shunt their firing) has been shown to prevent seizures, supporting a cell-autonomous mechanism30. This discrepancy may be reconciled by the identification of abnormal HCN4 channel expression in these neurons, which contributed to increased excitability by depolarizing RMPs, i.e., bringing the neurons closer to the AP threshold, and conferring firing sensitivity to intracellular cyclic AMP30. The abnormal expression of HCN4 channels in the RhebS16H neurons was found to be mTORC1-dependent by their sensitivity to rapamycin and the expression of constitutive active 4E-BP1. In addition, the abnormal HCN4 channel expression has been verified in resected human FCDII and HME tissue (of unknown genetic etiology)30,31. Here, we found that RhebY35L, mTORS2215Y, Depdc5KO, PtenKO, and Tsc1KO cortical neurons also display aberrant HCN4 channel expression, which is consistent with the mTORC1-dependent mode of expression in RhebS16H neurons30,31. The functional expression of these channels was variable, and the corresponding changes in Ih did not reach statistical significance for RhebY35L, Depdc5KO, and PtenKOneurons. Nonetheless, consistent with the function of HCN4 channels in maintaining the RMP at depolarized levels38, RhebY35L, mTORS2215Y, and Tsc1KO neurons exhibited depolarized RMPs. Interestingly, the RMP was unchanged in PtenKO and Depdc5KO neurons. The lack of RMP changes may be explained by the milder HCN4 phenotype or the presence of different ion channel complements in these neurons. In addition to HCN4 channels, it is thought that increased Kir channels, which are well-known to hyperpolarize RMPs, contribute to the enlarged inward currents in all the conditions we investigated. This was confirmed in the Tsc1KO neurons where application of BaCl2 to block Kir channels reduced the overall inward current and depolarized RMPs. Since PtenKO neurons have a larger overall inward current but a smaller Ih compared to Tsc1KO neurons, we postulate that PtenKO neurons have a higher expression of Kir channels neurons that could counteract the HCN4 channels’ depolarizing effect on RMP. The mechanisms and functional significance of the Kir channel increases in these neurons are yet to be validated. Furthermore, studies using rapamycin in each of the gene conditions would provide more direct evidence for the mTORC1-dependency of the HCN4 channel expression. Nevertheless, our data suggests that abnormal HCN4 channel expression is a conserved mechanism across these mTORC1-activating gene conditions and warrants further investigation of HCN-mediated excitability in mTORC1-related epilepsy.

While membrane excitability was largely similar across all gene conditions, the excitatory synaptic properties were more variable. All conditions, except for RhebY35L, led to increased excitatory synaptic activity, with variable impact on sEPSC frequency and amplitude. Tsc1KO neurons were the only investigated condition that displayed increased sEPSC frequency. This finding corroborates a previous study showing increased sEPSC frequency with no sEPSC amplitude changes in layer 2/3 cortical pyramidal neurons from 4-week-old Tsc1 conditional KO mice58. Interestingly, as previously reported for RhebS16H neurons59, the sEPSC frequency in RhebY35L neurons was reduced by 25% but this did not reach statistical significance, possibly due to a generally weaker effect of the Y35L mutation compared to the S16H mutation. RhebY35L, mTORS2215Y, and PtenKO neurons exhibited increased sEPSC amplitudes. We found no changes in the sEPSC amplitude of Depdc5KOneurons, which differs from a previous study reporting increased sEPSC amplitude in these neurons26. This discrepancy may likely be explained by the differences in the age of the animals examined in their study (P20-24) versus ours (P26-43), given that dynamic changes in synaptic properties are still ongoing past P2160. Despite the above differences in sEPSC frequencies and amplitudes, the total charge was increased in almost all conditions except for in RhebY35L, suggesting enhanced synaptic drive and connectivity. These excitatory synaptic changes may counteract the increases in rheobase (i.e., decreased membrane excitability) and thereby impact circuit hyperexcitability and seizure vulnerability. Overall, the changes in synaptic activity for the RhebY35L condition are in stark contrast to the other gene conditions. These observations suggest a Rheb-specific impact on synaptic activity that differs from the other mTORC1 pathway genes. Thus, despite impinging on mTORC1 signaling, different mTORC1 pathway gene mutations can affect synaptic activity, and thereby, excitability differently. The mechanisms accounting for the observed differences are not clear, but it is increasingly acknowledged that additional pathways beyond mTORC1 are activated or inactivated by each gene condition that could potentially contribute to the differential impacts on synaptic activity13. Although all the examined gene conditions activate mTORC1 signaling, they differentially impact the mTORC2 pathway. mTORC2 is a lesser understood, acute rapamycin-insensitive complex formed by mTOR binding to Rictor and known to regulate distinct cellular functions, including actin cytoskeletal organization10. Loss of PTEN increases mTORC2 activity, whereas loss of TSC1/2 and DEPDC5 is associated with decreased mTORC2 activity61–65. mTORS2215Y and RhebY35Lare not expected to change mTORC2 activity, although this remains to be verified. Studies have shown that inhibition of mTORC2 reduces neuronal overgrowth but not the synaptic defects and seizures associated with PTEN loss66,67. However, other studies have reported that inhibition of mTORC2 reduces seizures in several epilepsy models, including the mTORS2215F gain-of-function and Pten KO models68,69. Thus, the contribution of mTORC2 in epileptogenesis and seizure generation remains unclear, and future studies aimed to address the contribution of mTORC2 to the neuronal properties and synaptic activity in the different gene conditions are important to pursue.

Electrophysiological data from cytomegalic pyramidal neurons in cortical tissue obtained from humans with FCDII or TSC and intractable seizures have been reported70–73. These studies showed that human cytomegalic neurons have increased capacitance and decreased input resistance71,72, consistent with our findings in the present and previous studies in mice30,31. Further comparisons between pyramidal neurons in human TSC and FCDII cases showed similar changes in passive membrane and firing properties but differences in sEPSCs properties between the two conditions73. In particular, the frequency of sEPSC was higher in neurons in TSC compared to FCDII. These data were similar to our findings showing increased sEPSC frequency in the Tsc1KO condition but not in the other key gene conditions associated with FCDII. The authors of the 2010 human electrophysiology study concluded that although TSC and FCDII share several histopathologic similarities, there are subtle functional differences between these disorders73, aligning with the overall conclusions from our study. However, because the genetic etiology of the FCDII cases in these human studies is unknown, it is not possible to fully compare our gene-specific data to information published in these studies. Another study has reported functional reduction of GABAA-mediated synaptic transmission in cortical pyramidal neurons in individuals with FCDII74. In our study, we did not examine inhibitory synaptic properties and the impact on excitatory-inhibitory balance; this would be an important subject to pursue in another study.

Brain somatic mutations causing FMCDs, such as FCDII and HME, occur throughout cortical neurogenesis in neural progenitor cells that give rise to excitatory (pyramidal) neurons8,9. In the present study, we performed electroporation at E15, which targets progenitor cells that generate layer 2/3 pyramidal neurons in mice. Given that somatic mutations in FMCDs may occur at various timepoints during brain development, it would be interesting to examine the effects of mTOR pathway mutations in other cortical neuronal populations. For example, it would be interesting to investigate whether targeting earlier-born, layer 5 neurons by electroporating at E13 would result in similar or distinct phenotypes compared to our present observations in layer 2/3 neurons, and whether this would mitigate or accentuate the differences between the gene conditions. These findings would provide further insights into somatic mutations and mechanisms of FMCD and epilepsy.

In summary, we have shown that mutations affecting different mTORC1 pathway genes have similar and dissimilar consequences on cortical pyramidal neuron development and function, which may affect how neurons behave in cortical circuits. Our findings suggest that cortical neurons harboring different mTORC1 pathway gene mutations may differentially affect how neurons receive and process cortical inputs, which have implications for the mechanisms of cortical hyperexcitability and seizures in FMCDs, and potentially affect how neurons and their networks respond to therapeutic intervention.